Note on backward compatibility

A user may easily get a new container image that requires framework
script support not present in the user's environment, e.g., has
not yet done an update-labtainer.sh.  

A user may have an old container not supported by framework scripts
obtained via an update-labtainer.sh

\section{Lab Versions}
Substantive changes to an existing published lab should be made in
a new named lab.  A "substantive" change is defined as one that would
break any existing installation.  The sole exception is the case in 
which a newer lab image would require a Labtainers update.

Never change container names for existing labs.

Management of multiple versions of the same lab will be achieved entirely
by the user environment and what is distributed to it.

\subsection{Managing versions}
Labtainer updates will not remove old labs from installed systems, however 
they may stop distributing old labs.

If a user installation contains two versions of a lab, only the latest will
be displayed in the labtainer command -- unless the installation includes
containers for the older -- in which case both will be listed.

If a user starts an older version of a lab for the first time, warn the user
with a prompt.  Subsequently, note the existence of the newer lab without a prompt.

Lab names change with versions, e.g., centos-logv2.  (tab completion would really
help here).

Defining multiple labs as being different versions of the same lab is achieved
in a "version" file in the lab/config directory.  Absence of the file implies the
lab is its only version and is its base name. Otherwise, the file includes the 
lab basename and its version number, which is syntatically independent of the name
of the lab, e.g., lab names need not incorporate a "v2".  Version numbers are integers.
Lab basnames are ONLY used to distinguish versions of the same lab.   Students and
instructors name labs using their lab name, not their basename.

\subsection{Images dependent on framework versions}
The sole exception to the requirement for compatability of lab versions is the
case where we can request the user to update the labtainers framework in order
to run the new labs.  This concept is not limited to versions of the same lab,
it also pertains to changes to scripts in images that require corresponding
support from the framework.  An example might be a new argument to the
parameterize.sh script.  If new images expect this parameter, then the framework
must provide it, i.e., it must be updated.  (Backward compatability must still be
supported, i.e., old images MUST run with all new frameworks.)

These versions relationships will be managed crudely.  Once any new lab requires an
updated framework, all subsequent labs will require at least that framework version,
(whether it needs it or not).

These versions are managed by a version number imbedded in the build process.
