\input{header}
\begin{document}

\begin{center}
{\LARGE Buffer Overflow: 64 Bit Application}
\vspace{0.1in}\\
\end{center}

\section{Overview}
The {\tt bufoverflow} lab introduced you to buffer overflow vulnerabilities and potential exploits of those
vulnerabilities.  That lab included a vulnerable program that ran as a 32-bit x86 application.  
This lab includes the very same vulnerable program source code, however it compiles and runs as a 64-bit application.

\subsection {Background}
The student is expected to have an understanding of the Linux command line, and some amount of low level programming.
It is expected that the student will have completed the {\tt bufoverflow} lab.

\section{Lab Environment}
\label{environment}
This lab runs in the Labtainer framework,
available at http://nps.edu/web/c3o/labtainers.
That site includes links to a pre-built virtual machine
that has Labtainers installed, however Labtainers can
be run on any Linux host that supports Docker containers.

From your labtainer-student directory start the lab using:
\begin{verbatim}
    labtainer buf64
\end{verbatim}
\noindent A link to this lab manual will be displayed.  

The home directory of the resulting computer contains the source code of the vulnerable program ({\tt stack.c}) and
a template for the program that constructs the malicious data file ({\tt exploit.c}).  It also includes files for generating
shell code.  

\section{Tasks}
You will modify the {\tt exploit.c} program such that it generates a malicious data file that will cause the
vulnerable program to enter a shell.  This lab does not require that you get a root shell -- an application shell
is sufficient.  And your exploit need only run with Address Space Layout Randomization disabled, with an executable
stack, and with stack protection disabled.

A learning objective of this lab is to appreciate some of the differences between 32-bit and 64-bit x86 applications,
and how those differences might affect vulnerabilities and exploits.


\subsection{Explore}
Review the differences between the files in this lab, and the files in the {\tt bufoverflow} lab.  Note the {\tt stack.c}
files are the same -- but with potentially different buffer sizes.  Look at the assembly in {\tt shell.c} and compare that
to the assembly comments to the object code found in the {\tt bufoverflow exploit.c} file.

Disable ASLR:
\begin{verbatim}
  sudo sysctl -w kernel.randomize_va_space=0
\end{verbatim}
\noindent and use the {\tt compile.sh} script to compile the C programs and assemble the shell.asm
Run the {\tt stack} program.  Run it in the debugger.  Explore.

\subsection{Shell code}
In this lab, you will need to update the {\tt exploit.c} program to include the shell code.  Observe that the 64-bit
shell code has been assembled into the shell.bin file.  You must figure out how to get that into your {\tt exploit.c}
program.  Note you have been provided with a Python script called {\tt hexit.py}, and that may be of use.  Look at the
file from the {\tt bufoverflow} lab as an example.

\subsection{Overwriting return address}
By now you should have observed that the {\tt stack} program crashes when it encounters a bad return address.
As in the {\tt bufoverflow} lab, you will want to control that return address value.

What might you want that return address value to be?  Consider the properties of such a value and how that might
affect the {\tt strcpy} function behavior.

\subsection{Get a shell}
Alter the {\tt exploit.c} so that it generates a badfile that causes the {\tt stack} application to give you  a shell.
Once you get a shell, cat the exploit.c file from within the shell:
\begin{verbatim}
   cat exploit.c
\end{verbatim}

\subsection{Follow on}
Think about how you might approach an exploit if the {\tt stack.c} program were compiled without disabling stack protection
and stack execution.

\section{Submission}
After finishing the lab, go to the terminal on your Linux system that was used to start the lab and type:
\begin{verbatim}
    stoplab 
\end{verbatim}
When you stop the lab, the system will display a path to the zipped lab results on your Linux system.  Provide that file to 
your instructor, e.g., via the Sakai site.

\copyrightnotice

\end{document}
